---
title: 'ETM 540 Group Project'
author:
- Samira Akther  
- Jordan Hilton
- Andey Nunes
- Aparna Raghuram
- Mohammed Sheikh
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---
```{r setup, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library(magrittr, quietly = TRUE) #Used for pipes/dplyr
#library(dplyr, quietly = TRUE)
library(tidyverse, quietly = TRUE)
library(ROI, quietly = TRUE)
library(ROI.plugin.glpk, quietly = TRUE)
library(ompr, quietly = TRUE)
library(ompr.roi, quietly = TRUE)
library(pander, quietly = TRUE)
library(Benchmarking, quietly = TRUE)
library(rDEA, quietly = TRUE)
#library(reshape2, quietly = TRUE)
library(MultiplierDEA)
library(ggthemes)
library(corrplot)
#library(ggplotify)
#library(plotly)
```

# Efficiency of Public Expenditure on Education

In this section, the paper aims to study the efficiency of the public expenditures on education of OECD (Organization for Economic Development) countries. DEA (Data Envelopement Analysis), which is a vastly used tool for measuring efficiency, has been used to measure the efficiency of the public expenditures on education by the countries. The analysis provides an efficient frontier which is a linear combination of the data point with efficiency scores of 1. The analysis also allows us to know possible reduction in the inputs to generate the output of the countries with scores less than 1. 

## Optimization Model Definition

In our paper, we used output oriented model. Here, only one input is considered and defined as $x_{i,j}$ to  be the amount of the _i_'th input used by unit _j_ . Here j is the country we are analysing. Two outputs have been considered and defined as  $y_{r,j}$ to be the amount of the _r_'th output produced by unit _j_. n is the number of countries we analysed.  

The following optimization model has been solved to obtain the efficiency score while minimizing the value of $$\theta$$.


$$
 \begin{split}
 \begin{aligned}
 \text{Min  }   & \theta\\
  \text{subject to }\ & \sum_{j=1}^{n} x_{i,j}\lambda_j \leq x_{i,k} \forall \; i & i=1 ,\ldots, n\\
                      & \sum_{j=1}^{N^D} y_{r,j}\lambda_j \geq  y_{r,k} \forall \; r & j=1 ,\ldots, r&\\
                      & \lambda_j \geq 0  \; \forall \; j
  \end{aligned}
  \end{split}
$$

Here, Theta provides the efficiency scores of the countries we analyzed. The country with score of 1 lies in the efficient frontier and shows that any input reduction is not possible for generating the output. The countries with scores less than 1 will be considered as inefficient within the efficient frontier. Vector Lambda he specific amount of a unit _j_ used in setting the target for for performance for unit _k_.

## Data Description

Data has been obtained from secondary sources. Data on PISA scores were obtained from OECD website. Data on public education expenditures as percentage of GDP were obtained from UNESCO website. We considered 36 countries of OECD. The input variable is the public education spending as percentage of GDP of all 36 countries for the year of 2015. The output variables are the average PISA score of all 36 countries and the graduation rates for the year of 2015. 

Here is a summary of the data.


```{r load}
## load our data, downloaded from the google sheet
data <- read.csv("OECDdata.csv", stringsAsFactors = F)

glimpse(data)
```

We can see that there are three observations missing from several of the population education level fields and three zero values for `Total Spending`. These countries (listed below) will be excluded from the analysis.


```{r data cleaning}

data$Total.Spending <- as.numeric(gsub(",", "", data$Total.Spending))


# remove 0 spending countries
DEAdata <- data %>%
   filter(Total.Spending > 0)

setdiff(data[,1], DEAdata[,1])

```



# DEA Model
## Part A: Single input, single output

For this first part, we will set up and solve a DEA model using the single GDP data point for the input and the aggregated test score average for the single output.

Run a DEA with a single input (total education spending) and a single output (Average PISA score)
```{r run single dea}
#this is single input and single output
x <- DEAdata %>% select(Total.Spending) ## input
#x2 <- data %>% select(Spending.as.percentage.of.GDP) ## ratio input used in next section
y <- DEAdata %>% select(AveragePISA) ## output
row.names(x) <- DEAdata$Countries #%>%
   
row.names(y) <- DEAdata$Countries


ressingle <- DeaMultiplierModel(x, y, rts = "vrs", orientation = "output")

efficiencyTable <- tibble(
   country = as.character(dimnames(ressingle$Efficiency)[[1]]),
   efficiency = as.numeric(ressingle$Efficiency)) %>%
   arrange(desc(efficiency))

pander(head(efficiencyTable), caption = "DEA Output Efficiency for Aggregate Scores")
pander(head(ressingle$Lambda), caption = "DEA lambda values for Aggregate Scores")

```



```{r single output plot}
dea.plot(x = x, y = y, txt = dimnames(x)[[1]], GRID = T,
         xlab = "Total Spending", ylab = "Average PISA",
         ylim = c(400,550), main = "2015 OECD data DEA plot")
```



```{r hand drawn frontier single input}
# warning, the geom_segment calls are not reproducible and must be hand specified
ggplot(DEAdata, aes(x = Total.Spending, y  = AveragePISA)) +
   geom_point() +
   geom_text(aes(label = Countries), size = 3,
             nudge_x = 2, nudge_y = -1.5, color = "gray40") +
   geom_segment(aes(x = 93000, y = 529, xend = 39541, yend = 528.67), color = "dodgerblue") + # end of graph to Japan
   geom_segment(aes(x = 39541, y = 528.67, xend = 22149, yend = 523.67), color = "dodgerblue") + # from Japan to Canada
   geom_segment(aes(x = 22149, y = 523.67, xend = 16691, yend = 458.67), color = "dodgerblue") + # from Canada to Greece
   geom_segment(aes(x = 16691, y = 458.67, xend = 14173, yend = 415.67), color = "dodgerblue") + # from Greece to Mexico
   geom_segment(aes(x = 14173, y = 415.67, xend = 14173, yend = 400), color = "dodgerblue") + # Mexico to floor
   theme_minimal() +
   ggtitle("DEA plot of 2015 OECD Countries",
           subtitle = "Average Education Scores by Total Spending")
ggsave("single output DEA.png")
```

## Part B: 
Run a DEA with multiple inputs, and multiple outputs:

```{r runmultipledea}
#this is single input and multiple output
y <- DEAdata %>% select(SciencePISA, ReadingPISA, MathematicsPISA) ## output
row.names(y) <- DEAdata[,1]

resmult <- DeaMultiplierModel(x, y, rts = "vrs", orientation = "output")

multEfficiencyTable <- tibble(
   country = as.character(dimnames(resmult$Efficiency)[[1]]),
   efficiency = as.numeric(resmult$Efficiency)) %>%
   arrange(desc(efficiency))
```


```{r inspect resmult}
pander(head(multEfficiencyTable, 10), caption = "DEA Output Efficency for Multiple Scores")
pander(head(resmult$Lambda), caption = "DEA lambda values for Multiple Scores")
```


```{r multiple output plot}
dea.plot(x = x, y = y, txt = dimnames(x)[[1]], GRID = T,
         xlab = "Total Spending", ylab = "Multiple PISA",
         ylim = c(1200,1600), main = "2015 OECD data DEA plot")
```


```{r hand drawn multiple output frontier}
# warning, the geom_segment calls are not reproducible and must be hand specified
DEAdata$multPISA <- DEAdata$SciencePISA + DEAdata$ReadingPISA + DEAdata$MathematicsPISA
DEAdata$country <- DEAdata$Countries

ggplot(DEAdata, aes(x = Total.Spending, y  = multPISA)) +
   geom_point() +
   geom_text(aes(label = Countries), size = 3,
             nudge_x = 2, nudge_y = -1.5, color = "gray40") +
   geom_segment(aes(x = 93000, y = 1590, xend = 39541, yend = 1586), color = "dodgerblue") + # end of graph to Japan
   geom_segment(aes(x = 39541, y = 1586, xend = 22149, yend = 1571), color = "dodgerblue") + # from Japan to Canada
   geom_segment(aes(x = 22149, y = 1571, xend = 16691, yend = 1376), color = "dodgerblue") + # from Canada to Greece
   geom_segment(aes(x = 16691, y = 1376, xend = 14173, yend = 1247), color = "dodgerblue") + # from Greece to Mexico
   geom_segment(aes(x = 14173, y = 1247, xend = 14173, yend = 1200), color = "dodgerblue") + # 
   theme_minimal() +
   ggtitle("DEA plot of 2015 OECD Countries",
           subtitle = "Multiple Education Scores by Total Spending")
# ggsave("multiple output DEA.png")
```



## Part C: Results and discussion

Some quick analysis of the results:

```{r quickanalysis}
max(ressingle$Efficiency) ## the maximum efficiency for one output
which.max(ressingle$Efficiency) ## the row value of the maximum efficiency
data[33, ] ## the row from the data matching the answers above
min(ressingle$Efficiency) ##same for mins
which.min(ressingle$Efficiency)
data[22,]
ressingle$Efficiency["United States",] ## US efficiency

max(resmult$Efficiency) ## the maximum efficiency for one output
which.max(resmult$Efficiency) ## the row value of the maximum efficiency
data[4, ]
min(resmult$Efficiency) ##same for mins
which.min(resmult$Efficiency)
data[14,]
```

example change

# References

### Data Sources

> insert links/doi to data sets/info here

Anderson, T. R. (2019) Operations Research in R 

Anderson, T. R. (2019) Data Envelope Analysis in R

Bogetoft, P., Otto, L. (2011) Benchmarking with DEA, SFS, and R. Springer. 

# Appendix

