---
title: "Project Linear Program"
author:
- Samira Akther  
- Jordan Hilton
- Andey Nunes
- Aparna Gandikota
- Mohammed Sheikh
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library(magrittr, quietly = TRUE) #Used for pipes/dplyr
#library(dplyr, quietly = TRUE)
library(tidyverse, quietly = TRUE)
library(knitr, quietly = TRUE)
library(ROI, quietly = TRUE)
library(ROI.plugin.glpk, quietly = TRUE)
library(ompr, quietly = TRUE)
library(ompr.roi, quietly = TRUE)
library(pander, quietly = TRUE)
```
Let's load our data and apply some variable names. We're going to use Average PISA score as our desired output and total spending per student as the input. We also need to drop some rows for this since they are missing the "total education spending" variable we want to use:


```{r load}
## load our data, downloaded from the google sheet
eddata <- read.csv("OECDdata.csv")


eddata <- eddata[-c(7,21,33),] ##drop rows missing data
countrynames <- eddata[,1]
AvPisa <- as.numeric(eddata[,6]) ## average PISA score
TotSpending <- as.numeric(gsub(",", "", eddata[,11])) ## total spending per student in each country
SpendEff <- AvPisa/TotSpending #the spending efficience per test score in each country
countrycount <- length(countrynames) #the total number of countries
```

For our linear program, we're going to pretend for a moment that all these countries have equal student populations,  and that if we look at the current ratio of spending per student and test score achievement, that will tell us how efficient it is to spend additional money in each country.

Here's our question for analysis: let's say we have a grant of size $\$5,000$ student, and that we can't spend more than  $\$1,000$ student in any single country. How should we spend our grant to most efficiently improve test scores?

We have 36 countries in the sample, so we have 36 decision variables $x_i$; how much money to spend in each country. We should not spend less than $\$0$ in any country.

For this formulation let's define the decison variable as $x_i$ and the spending efficiency as $E_i$:

$$
 \begin{split}
 \begin{aligned}
    \text{Maximize  }   &  \sum_{i=1}^{\#countries} E_i * x_i \\
    \text{subject to }  & \sum_{i=1}^{\#countries} x_i \leq 5000\\
                        & x_i \leq 1000, \\
                        & x_i \geq 0  \;  \forall \; i
  \end{aligned}
  \end{split}
$$


```{r impliciteducationLP}
educationLP <- MIPModel() %>%
  add_variable(x[i], i = 1:countrycount, type = "continuous", lb = 0) %>%
  set_objective(sum_expr(SpendEff[i] * x[i] , i = 1:countrycount), "max") %>% #maximize test scores
  add_constraint(sum_expr(x[i], i = 1:countrycount) <= 5000) %>%  # total spending constraint
  add_constraint(x[i] <= 1000,  i = 1:countrycount) %>%  # per country spending restraint
  solve_model(with_ROI(solver = "glpk"))
  
educationLP
```

Let's take a look at our solution:

```{r inspect}
edLPsolution <- educationLP$solution
names(edLPsolution) <- countrynames
kable(edLPsolution)
```



```{r solution countries}
solutioncountries <- as.character(countrynames[c(5,11,12,21,31)])
solutionscores <- AvPisa[c(5,11,12,21,31)]
solutionspending <- as.numeric(TotSpending[c(5,11,12,21,31)])
pander(cbind(solutioncountries, solutionscores,solutionspending))
```



#### Shadow Prices **Row Duals**

```{r shadow prices}
shadow_prices <- educationLP$solution_row_duals()
names(shadow_prices) <- countrynames
shadow_prices
```


#### Reduced Costs **Column Duals**

```{r reduced costs}
reduced_costs <- educationLP$solution_column_duals()
names(reduced_costs) <- countrynames
kable(reduced_costs)
```





